Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-28 manuscript/manuscript.tex
Add citation to Lin2018 in Appendix
Edits to the density discretization and software
Minor edits to the algorithm to make it more like the adaptive
discretization description.
Added some more comments about how we can use any function because the
GLQ is the same and our software can handle user defined functions.

The numerical integration of an arbitrary continuous density function introduces a new
type of problem: the integration error from using only a few nodes to account for the
variation of the density function.
The three dimensional adaptive discretization may help
to reduce this kind of error by adding more point masses in the radial direction.
However, it does not take into account the density function when subdividing the
tesseroid and hence it is not well suited to fully perform this task.
The two dimensional adaptive discretization algorithm does not subdivide the tesseroid
on the radial dimension, so it does not put any effort on lowering this new kind of
error.

We have developed a complementary discretization
algorithm that takes into account the variations of the density function.
This density-based discretization is applied prior to the adaptive discretization
described in the previous section.
In short, the algorithm divides the tesseroid along the radial dimension at the
depths at which the \emph{maximum density variations} take place.

Consider an \emph{original} tesseroid with a continuous density given by the function
$\rho(r')$.
Before the density-based discretization starts,
we normalise the density function to the range $[0, 1]$ as follows

@@

\textit{Step 1}:
Define a linear function $\rho_l(r')$ that assumes
the same values as the normalised density $\rho_n(r')$ at the boundaries of the
tesseroid ($r_1$ and $r_2$):

@@

\textit{Step 5}:
If inequality~\ref{eq:delta-density} is not satisfied, then the tesseroid is split in
two parts at the radius $r_\text{max}$ at which the maximum absolute difference
(Eq.~\ref{eq:density-abs-diff}) takes place.

\textit{Step 6}:
Repeat steps 1-6 for each smaller tesseroid produced in step 5.

Once all smaller tesseroids satisfy Eq.~\ref{eq:delta-density},
each one is subjected to the two dimensional adaptive
discretization algorithm described earlier to calculate their gravitational effects.

@@

In summary, given any tesseroid with variable density in depth and an arbitrary
external computation point, we propose the following steps to numerically
compute the gravitational fields:

\textit{Step 1:}
Apply the density-based discretization algorithm.
It subdivides the tesseroid only in the radial dimension, thus it will return a set of
tesseroids with the same longitudinal and latitudinal dimensions as the original one but
with different radial boundaries.

\textit{Step 2:}
Apply the two dimensional adaptive discretization algorithm on each tesseroid obtained
in the previous step.
If needed, it will subdivide each tesseroid only in the latitudinal and longitudinal
direction, generating a set of smaller tesseroids that resemble the original one.

\textit{Step 3:}
Apply a second-order GLQ to numerically compute the gravitational fields
(Eq.\ref{eq:glq-var-dens}) generated by each tesseroid obtained in the
previous step.
Then sum these results in order to obtain the gravitational field of the original
tesseroid.

@@

We have implemented the algorithms described in the previous sections in the Python
programming language.
The software is based on the pre-existing code for the homogeneous density tesseroid
\citep{Uieda2016}, more specifically the implementation in the Python library Fatiando a
Terra v0.5 \citep{Uieda2013}.
The more time consuming parts of the algorithm are written in the Cython language to
achieve higher performance.
This new code is freely available under the BSD 3-clause open-source license.
It can be downloaded from the online repository
\href{https://github.com/pinga-lab/tesseroid-variable-density}{github.com/pinga-lab/tesseroid-variable-density}.

>>>>

The numerical integration of an arbitrary continuous density function introduces a new
type of problem: the integration error from using only a few quadrature nodes to account
for the variation of the density function.
The three dimensional adaptive discretization may help to reduce this kind of error by
adding more point masses in the radial direction.
However, it does not take into account the density function itself when dividing the
tesseroid and hence it is not well suited to fully perform this task.

We have developed a complementary discretization algorithm in the redial direction that
takes into account the variations of the density function.
This density-based discretization is applied prior to the two dimensional adaptive
discretization described in the previous section.
In short, the algorithm divides the tesseroid along the radial dimension at the
depths at which the \emph{maximum density variations} take place.

Consider an \emph{original} tesseroid with a density given by the continuous function
$\rho(r')$.
Before the density-based discretization starts,
we normalise the density function to the range $[0, 1]$ as follows

@@

\textit{Step 1}:
Define a linear function $\rho_l(r')$ that assumes the same values as the normalised
density function $\rho_n(r')$ at the boundaries of the tesseroid ($r_1$ and $r_2$):

@@

\textit{Step 5}:
If inequality~\ref{eq:delta-density} is not satisfied, then the tesseroid is split in
two parts at the radius $r_\text{max}$ at which the maximum absolute difference
(Eq.~\ref{eq:density-abs-diff}) takes place.
Repeat steps 1-5 for each smaller tesseroid produced in this step.

\textit{Final step}:
Once all smaller tesseroids satisfy inequality.~\ref{eq:delta-density}, each one is
subjected to the two dimensional adaptive discretization algorithm described earlier to
calculate their gravitational effects.

@@

In summary, given a tesseroid with density varying in depth according to an arbitrary
continuous function, we propose the following steps to numerically compute its
gravitational fields on an external point:

\textit{Step 1:}
Apply the density-based discretization algorithm to discretize the tesseroid in the
radial dimension, producing a set of tesseroids with the same longitudinal and
latitudinal dimensions as the original one but with different radial boundaries.

\textit{Step 2:}
Apply the two dimensional adaptive discretization algorithm for each tesseroid obtained
in the previous step.
If needed, the algorithm will divide each tesseroid in the latitudinal and longitudinal
direction, generating a set of smaller tesseroids.

\textit{Step 3:}
Apply a second-order GLQ to numerically compute the gravitational fields
(Eq.\ref{eq:glq-var-dens}) generated by each tesseroid obtained in the
previous step.
The numerical integration includes the density function and can be applied
without modification to any continuous function.
The sum of these results is the gravitational field of the original tesseroid.

@@

We have implemented the algorithms described in the previous sections in the Python
programming language.
The software is based on the Python implementation of \citet{Uieda2016} that is included
in the Fatiando a Terra library v0.5 \citep{Uieda2013}.
The more time consuming parts of the algorithm are written in the Cython language to
achieve higher performance.
We leverage the dynamic nature of the Python language to allow user-defined radial
density functions as inputs to the software.
Thus, our code can evaluate linear, exponential, polynomial, sinusoidal, cubic splines,
or any other continuous density function without modification.
This new code is freely available under the BSD 3-clause open-source license.
It can be freely downloaded from the online repository
\href{https://github.com/pinga-lab/tesseroid-variable-density}{github.com/pinga-lab/tesseroid-variable-density}.
